### 2023-03-09 Manuscript/Draft1.Rmd
manuscript update

The performance of our abundance estimates tend to decrease under increasing temperature, with the most dramatic changes seen in our VAST estimates without covariates. One exception to this is when we use a stratified mean approach while the population is decreasing. Since our analyses have found that the stratified mean tends to under estimate the true abundance, as the Yellowtail population decreases towards zero the difference between the estimate and the true value also decrease. That is, if the population is low enough, failing to appropriately sample the population in a design-based method produces the same result as appropriately sampling.

@@

As seen in Figure \@ref(fig:PopPct), in the constant temperature scenario the Cod population is decreasing its presence in strata 19 and 20 over the duration of the simulation in both seasons, while simultaneously increasing presence in stratum 16. However, an increase in average temperature had a seasonal impact on Cod's presence in strata 16 in our simulations. In the spring the population followed a similar trend of leaving 19 and 20 while simultaneously entering 16, but in the fall the increasing temperature drives the population out of strata 16 and 21, and into 18, 22, and 24. **check the impact of this difference on estimates**

Of the abundance estimates without covariates, $12/16=75$\% had a VAST fit with a lower relative error compared to the stratified mean estimate. Including detailed covariate information in our Cod VAST estimates resulted in all $16/16=100\%$ having a lower relative error than the corresponding stratified mean estimate. The 4 scenarios without covariates that had a higher VAST error were each overestimates for the fall season that involved an increasing temperature gradient. As seen in Figure XXX, Cod are shifting their spatial distribution much more in the fall than they are in the spring. While adding covariates to these fits allowed for an improved error relative to the stratified mean, the VAST abundance still display an overestimate early in the time series before covariates correct the trend, allowing for a lower overall absolute error **(show these plots? IncTemp fall scenarios)**.

@@

Figure \@ref(fig:PopPct) reveals some subtle seasonal differences in the percent of haddock in each strata. In the constant temperature scenarios, the spring shows a decrease in strata 19 and 20 that correspond to increases in 16, 29 and 24. While strata 24 and 29 also increase in the fall under constant temperature, the corresponding decrease is primarily from strata 16 and 13. While similar results can be seen in the spring for the increasing temperature scenario, much more dramatic results exist in the fall under increasing temperature as we see a significant decreases in strata 13, 16, 21, and 22 that leads to the most noticable increases in strata 17, 18, and 29. This shift represents movement from the less deep and more centrally located strata towards exterior deeper strata.

>>>>

The performance of our abundance estimates tend to decrease under increasing temperature, with the most dramatic changes seen in our VAST estimates without covariates. One exception to this is when we use a stratified mean approach while the population is decreasing. Our analyses have found that the stratified mean tends to under estimate the true abundance and since these estimates are bounded below by zero, as the Yellowtail population decreases towards zero the difference between the estimate and the true value also decrease. That is, if the population is low enough, failing to appropriately sample the population in a design-based method produces the same result as appropriately sampling.

@@

We see a seasonal impact of an increasing average temperature in stratum 16 in the Cod results in Figure \@ref(fig:PopPct). In the constant temperature scenario for Cod, in both seasons the population is decreasing its presence in strata 19 and 20 over the duration of the simulation, while simultaneously increasing presence in stratum 16. However, an increase in average temperature in the fall drives the population out of strata 16 and 21, and into 18, 22, and 24 (see Figure \@ref(fig:hab-plot)). Similar to the Yellowtail Flounder population, stratum the habitat in stratum 16 acts as an attractor in both temperature scenarios in the spring when the water temperature is cooler. When the temperature increases over time the fall population begins to prefer the adjacent strata that are deeper and/or more northern than stratum 16.

Of the abundance estimates without covariates, $12/16=75$\% had a VAST fit with a lower relative error compared to the stratified mean estimate. Including detailed covariate information in our Cod VAST estimates resulted in all $16/16=100\%$ having a lower relative error than the corresponding stratified mean estimate. The 4 scenarios without covariates that had a higher VAST error were each overestimates for the fall season that involved an increasing temperature gradient. Similar to Yellowtail Flounder results, Cod are shifting their spatial distribution into the smaller exterior strata in these conditions, which is decreasing the accuracy of the covariate-free VAST estimate. While adding covariates to these fits allowed for an improved error relative to the stratified mean, the VAST abundance still display an overestimate early in the time series before covariates correct the trend, allowing for a lower overall absolute error **(show these plots? IncTemp fall scenarios)**.

@@

Figure \@ref(fig:PopPct) reveals some subtle seasonal differences in the percent of haddock in each strata. In the constant temperature scenarios, the spring shows a decrease in strata 19 and 20 that correspond to increases in 16, 24 and 29. This change represents a northward movement between larger centrally located strata. While strata 24 and 29 also increase in the fall under constant temperature, the corresponding decrease is primarily from strata 16 and 13. While similar results can be seen in the spring for the increasing temperature scenario, much more dramatic results exist in the fall under increasing temperature as we see a significant decreases in strata 13, 16, 21, and 22 that leads to the most noticeable increases in strata 17, 18, and 29. This shift represents movement from the shallower and more centrally located strata towards exterior deeper strata. Since stratum 16 contains very favorable habitat including much of the species' spawning ground, the strong shift out of 16 and into the northern most strata of 29 in the fall demonstrates a climate-driven change in movement preference.

### 2023-03-09 Manuscript/Draft1.tex
manuscript update

As seen in Figure \ref{fig:PopPct}, in the constant temperature scenario the Cod population is decreasing its presence in strata 19 and 20 over the duration of the simulation in both seasons, while simultaneously increasing presence in stratum 16. However, an increase in average temperature had a seasonal impact on Cod's presence in strata 16 in our simulations. In the spring the population followed a similar trend of leaving 19 and 20 while simultaneously entering 16, but in the fall the increasing temperature drives the population out of strata 16 and 21, and into 18, 22, and 24. \textbf{check the impact of this difference on estimates}

Of the abundance estimates without covariates, \(12/16=75\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate. Including detailed covariate information in our Cod VAST estimates resulted in all \(16/16=100\%\) having a lower relative error than the corresponding stratified mean estimate. The 4 scenarios without covariates that had a higher VAST error were each overestimates for the fall season that involved an increasing temperature gradient. As seen in Figure XXX, Cod are shifting their spatial distribution much more in the fall than they are in the spring. While adding covariates to these fits allowed for an improved error relative to the stratified mean, the VAST abundance still display an overestimate early in the time series before covariates correct the trend, allowing for a lower overall absolute error \textbf{(show these plots? IncTemp fall scenarios)}.

@@

Figure \ref{fig:PopPct} reveals some subtle seasonal differences in the percent of haddock in each strata. In the constant temperature scenarios, the spring shows a decrease in strata 19 and 20 that correspond to increases in 16, 29 and 24. While strata 24 and 29 also increase in the fall under constant temperature, the corresponding decrease is primarily from strata 16 and 13. While similar results can be seen in the spring for the increasing temperature scenario, much more dramatic results exist in the fall under increasing temperature as we see a significant decreases in strata 13, 16, 21, and 22 that leads to the most noticable increases in strata 17, 18, and 29. This shift represents movement from the less deep and more centrally located strata towards exterior deeper strata.

>>>>

We see a seasonal impact of an increasing average temperature in stratum 16 in the Cod results in Figure \ref{fig:PopPct}. In the constant temperature scenario for Cod, in both seasons the population is decreasing its presence in strata 19 and 20 over the duration of the simulation, while simultaneously increasing presence in stratum 16. However, an increase in average temperature in the fall drives the population out of strata 16 and 21, and into 18, 22, and 24 (see Figure \ref{fig:hab-plot}). Similar to the Yellowtail Flounder population, stratum the habitat in stratum 16 acts as an attractor in both temperature scenarios in the spring when the water temperature is cooler. When the temperature increases over time the fall population begins to prefer the adjacent strata that are deeper and/or more northern than stratum 16.

Of the abundance estimates without covariates, \(12/16=75\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate. Including detailed covariate information in our Cod VAST estimates resulted in all \(16/16=100\%\) having a lower relative error than the corresponding stratified mean estimate. The 4 scenarios without covariates that had a higher VAST error were each overestimates for the fall season that involved an increasing temperature gradient. Similar to Yellowtail Flounder results, Cod are shifting their spatial distribution into the smaller exterior strata in these conditions, which is decreasing the accuracy of the covariate-free VAST estimate. While adding covariates to these fits allowed for an improved error relative to the stratified mean, the VAST abundance still display an overestimate early in the time series before covariates correct the trend, allowing for a lower overall absolute error \textbf{(show these plots? IncTemp fall scenarios)}.

@@

Figure \ref{fig:PopPct} reveals some subtle seasonal differences in the percent of haddock in each strata. In the constant temperature scenarios, the spring shows a decrease in strata 19 and 20 that correspond to increases in 16, 24 and 29. This change represents a northward movement between larger centrally located strata. While strata 24 and 29 also increase in the fall under constant temperature, the corresponding decrease is primarily from strata 16 and 13. While similar results can be seen in the spring for the increasing temperature scenario, much more dramatic results exist in the fall under increasing temperature as we see a significant decreases in strata 13, 16, 21, and 22 that leads to the most noticeable increases in strata 17, 18, and 29. This shift represents movement from the shallower and more centrally located strata towards exterior deeper strata. Since stratum 16 contains very favorable habitat including much of the species' spawning ground, the strong shift out of 16 and into the northern most strata of 29 in the fall demonstrates a climate-driven change in movement preference.
